---
title: "tree toy"
author: "Andrea Spano'"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

## Packages

```{r, include = TRUE, message=FALSE}
#ioslides_presentation:
require(dplyr)
require(ggplot2)
require(rpart)
require(rpart.plot)
```

## Basic 

In a basic scenario we have a sample of ten people. 
Each person has `(1)` or does not have `(0)` a given characteristic. 

```{r}
characteristic <-  c(1, 1, 1, 1, 0, 1, 1, 0, 0, 0)
characteristic
```

We wish to classify the whole sample as either `zero` or `one`; does the sample has the characteristic? 

As the majority of the set is made by characteristic of type `one`, we classify the whole sample as a sample of _class_ `one`.

The _Miss Classification Error_ (MCE) being:

```{r}
e0 <- 4/10
e0
```

as four characteristic, four zeros, are wrongly classified 




## Data

In a more complicated scenario we can have characteristic associated males of females

```{r}
data <- tibble(
    gender = rep(c('M', 'F'), each = 5), 
    characteristic = c(1, 1, 1, 1, 0, 1, 1, 0, 0, 0)

)
data
```


## Root classification {.columns-2 .smaller}


```{r}
data  |> 
    group_by(event)  |> 
    count()
```    

As in the root node `one` is the most frequent class, we classify the whole node as a node of class `one`

The miss classification error (MCE), where zero occurs, is therefore:

```{r}
e0 <- 4/10 
e0
```

## First split 

We now split our data based on variable `gender`

```{r}
split_gender <- data  |> 
    group_by(gender, event)  |> 
    summarise(n = n(), .groups = 'drop')
split_gender    
 ```   



## First split 

```{r}
split_gender
```

The `F` node is at prevalece `zero`. As a result, MCE is:

```{r}
e1_F <- 2 / (3+2); e1_F
```

The `M` node is at prevalece `one`. As a result, MCE is:

```{r}
e1_M <- 1 / (4+1); e1_M
```

## First split 

```{r}
split_gender
```

The over all MCE is:

```{r}
e1 <- (2+1) / (3+2+1+4); e1
```

## First split 

We rember the MCE at root node was 
```{r}
e0
```

Scaling on `e0`, `r0` becomes:

```{r}
r0 <- e0/e0
r0
```

while `r1` becomes:

```{r}
r1 <- e1/e0
r1
```


## toy tree

```{r}
fm <- rpart(event~gender, 
    data = data, 
    method = 'class',
    parms = list(split = 'gini'), 
    control = rpart.control(
        minbucket = 2,
        cp = 0.00001),
    xval = 2)
rpart.plot(fm)
```

## Cost of pruning 

```{r}
printcp(fm)    
```
```{r}
CP <- (1-0.75)/(1-0)
CP
```


